---
title: "ESP 106 Final Project"
author: Kyra Liu, Sung-A Lee
output: pdf_document
---

## Introduction

Despite being known for hot and dry weather conditions, the entire state of California is prone to floods; many of which can potentially cause widespread damage and disruption. Understanding the distribution of flood-risk zones and the type of land use in each zone is critical for developing effective flood management strategies.(Pattison & Lane, 2011). The state's land use patterns, particularly relating to rapid urbanization and agricultural development have altered natural drainage patterns, increasing the risk of flooding in areas that were previously less susceptible. In addition, development in flood-prone areas, such as near rivers or in low-lying regions, has put more people and infrastructure in harm's way.

Overall, California's complex combination of geography, climate, and land use makes it a challenging environment to manage flood risk (Huang et al., 2020). As local residents, we are acutely aware of the impact that flooding can have on Yolo and Sacramento County. In this project, we seek to delve deeper into the flood-risk zones and the types of land use in each zone. We aim to determine the distribution of flood risks across different land use types in Yolo and Sacramento County, California. Ulitmately we would like to employ this in developing strategies to better prepare for and mitigate the effects of future flooding events.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# knitr::opts_knit$set(root.dir= '/Users/sunga/Documents/GitHub/floodproject') # sung wd
# knitr::opts_knit$set(root.dir = "/Volumes/KYRADRIVE/floodproject") # kyra mac wd
library(terra)
library(ggplot2)
library(geodata)
library(maptiles)
library(wesanderson) # color palette

```

## Data Processing

This project employs three datasets. The first dataset is the National Flood Hazard Layer from the Federal Emergency Management Agency (FEMA). This is a geospatial database that contains current effective flood hazard data, primarily used to support the National Flood Insurance Program. To gain a better understanding of this data, we utilized the NFHL ArcGIS Viewer for visualization. After downloading the data and loading it into R, we were able to see the 32 layers present in the data. Initially, we extracted two layers, one containing flood-zone boundaries and one containing flood-zone areas. Taking the names of these two layers allowed us to understand that the flood-zone areas included the boundaries as well as the risk-types, leading us to focus on that layer for the remainder of our analysis. FEMA categorizes flood risks into 7 types. For the purposes of this project, the types were categorized as either high-risk, moderate-risk, or low-risk.

The second dataset is the 2019 statewide crop mapping GIS shapefile from the California Natural Resources Agency (CRNA) managed by the California Department of Water Resources. This dataset contains information about the distribution of crops as well as urban land in the state of California. This project was interested in only the differences between agricultural and urban land, so all of the crops were categorized as agricultural land.

Additionally, we obtained administrative boundaries for the U.S. at the county level using the geodata package in R. The package obtains data from the Database of Global Administrative Areas.

Much of our data analysis involved processing the data for use. Because of this, we created a function to help streamline the process, specifically for reproducibility. This function allows the user to put in any county in California and get land use data and national flood hazard layer for that county and saves it as rds. This allows us to process data for multiple counties much more efficiently. Also, it allows us to save a much smaller (cropped) data set for an individual county rather than having to access both statewide data sets all at once.

The function takes in a string of the county's name. If the county exists in California (case insensitive), then it crops both data sets to the specified county and saves each subset data as a RDS. If the county, as inputted, is not in California, the function breaks and returns an error which states that the county does not exist in California. Within this function, we obtain data for the USA from the geodata package in R and extract only border of the desired county. Then we use this to crop the flood and land-use data to only the input county. It tells the user when each file has been downloaded, or if the file already exists in the directory (in which case a new file is not created).

The function saves each landuse file in the format countynamelu.rds and each flood layer file as countynameflood.rds. The countyname here is the name of the specified county in all lowercase letters with no spaces. For example the files for for the input "Los Angeles" would be saved as losangeleslu.rds and losangelesflood.rds. If the desired RDS already exists, the function informs the user. Additionally, the function outputs a list of the data by reading the RDS file.

```{r}

flcounty <- function(countyname) {
  #if nfhl does not exist in directory, download from fema
countydata <- list(flood = NULL, landuse = NULL)
  clname <- tolower(countyname)
 cfiname <- gsub(" ", "", countyname) # files are saved without spaces in multiple word counties
    usa= geodata::gadm("USA", level=2, path=".")

  usa = usa[usa$NAME_1=="California"]
  county_idx <- grep(paste0(clname, collapse="|"), tolower(usa$NAME_2))
  county <- usa[county_idx, ]
  # stops if county is not in california
  if (length(county_idx) == 0) {
    stop(paste(countyname, "county not found in California"))
  } else if (length(county_idx) > 1) {
    warning(paste("Multiple matches found for county", countyname))
  }
  fldfiname <- "femadata/NFHL_06_20230220.zip"
  if (!file.exists(fldfiname)){
    dir.create("femadata")
    download.file("https://hazards.fema.gov/nfhlv2/output/State/NFHL_06_20230220.zip",
                  destfile= fldfiname,mode = "wb")
  }
  #file <- unzip(fldfiname, exdir = "./femadata")
  # take input county name as lower for string comparison
 
  # create name of rds to be created
  fldrds <- paste0(cfiname,"flood.rds")
  # if rds for nfhl county does not already exist in directory, create and save
  #checks is county is in california before cropping, also checks for duplicates
  if (!file.exists(fldrds)) {
   
    # subsets to county
   
    # creating spat vector from nfhl layer
    unzip("./femadata/NFHL_06_20230220.zip",exdir="./femadata")
    floodarea <- terra::vect("./femadata/NFHL_06_20230220.gdb",layer = "S_FLD_HAZ_AR")
    
    #cropping to county and saving to rds
    floodcounty = terra::crop(floodarea, county)
    saveRDS(floodcounty, fldrds)
    print(paste0(fldrds, " saved to directory"))
  }else{print(paste0(fldrds, " already exists in directory"))}
countydata$flood <- readRDS(fldrds)
  # if land use data does not exist in directory, download from CNRA
  landname <- "./landdata/i15_Crop_Mapping_2019.zip"
  if (!file.exists(landname)){
    dir.create("landdata")
    url2<- "https://data.cnra.ca.gov/dataset/6c3d65e3-35bb-49e1-a51e-49d5a2cf09a9/resource/1da7b37a-dd97-4b69-a86a-fe824a252eaf/download/i15_crop_mapping_2019.zip"
    download.file(url2,destfile= landname,mode = "wb")
    #unzip("./landdata/i15_Crop_Mapping_2019.zip")
    #landuse <- vect(usefiles[7])
  } 
  # create name of rds to be created
  lu_rds <- paste0(cfiname,"landu.rds")
  #if the rds for landuse does not exist for county, crops to county and saves
  if (!file.exists(lu_rds)) {
   
    unzip("./landdata/i15_Crop_Mapping_2019.zip", exdir= "./landdata")
    landuse <- terra::vect("./landdata/i15_Crop_Mapping_2019/i15_Crop_Mapping_2019.shp")
    lu_county = terra::crop(landuse, county)
    saveRDS(lu_county, lu_rds)
    print(paste0(lu_rds, " saved to directory"))
  }else{print(paste0(lu_rds, " already exists in directory"))}
  countydata$landuse <- readRDS(lu_rds)
  countydata
}

```

Using our created function, we obtained data for two counties, Yolo and Sacramento. We then joined them by appending the datasets so that we had one SpatVector for land use and one for flood risks. Then, we categorized the flood risks and the land use types for our project. In this project, we have 2 land-use classes: urban and agricultural, and 3 flood-risk categories: high, moderate, and low. We did this using several logical operators and subsetting the columns in order to change the data in that column to the desired categories. Once we categorized the data, we intersected the flood and land-use data. Using this intersected data, we created visualizations of the distribution of flood risk across the two counties for the different land-use types.

```{r}
# Using function to get data
yolodata <- flcounty('yolo')
floodyolo <- yolodata$flood # yolo flood risk data
yololu <- yolodata$landuse # yolo land use data
yololu$COUNTY <- "Yolo" # county as yolo

sacdata <- flcounty('sacramento')
floodsac <- sacdata$flood # sac flood data
saclu <- sacdata$landuse # sac land use data
saclu$COUNTY <- "Sacramento" # county as sacramento

# joining data for the two counties
landuse <- rbind(yololu,saclu) # land use for both counties
floodarea<- rbind(floodyolo,floodsac) # flood risk for both counties
initclasses <- unique(landuse$SYMB_CLASS)
knitr::kable(table(floodarea$FLD_ZONE),col.names = c("risk type","count"),
             caption= "Initial Counts for Flood Zone Designations in Data")
initzones <- unique(floodarea$FLD_ZONE)
knitr::kable(table(landuse$SYMB_CLASS),
             col.names = c('crop mapping category','count'),
             caption = "Initial Counts for Land-Use Designations in Data")
```

```{r}
# categorizing data
# assign agriculture area if the value is not equal to U 
agind <- landuse$SYMB_CLASS != "U" 
#urbanind<- grep(uses,cats[11],value = 0)
# replacing the values in the column
landuse$landuse <- "Urban"
landuse$landuse[agind]<- "Agricultural"

# categorizing flood risks
highcats <- c("A","AE","AH","AO","VE","V") # high risk flood zones defined by fema
modcats <- c("X") # moderate flood zones defined by fema
lowcats <- c("D","A99") # low risk flood zones defined by fema
floodarea <- floodarea[floodarea$FLD_ZONE != "AREA NOT INCLUDED" & floodarea$FLD_ZONE!= "OPEN WATER", ]
floodarea$zones <- "wrong"
#setting zones
floodarea$zones[floodarea$FLD_ZONE %in% highcats] <- "high risk"
floodarea$zones[floodarea$FLD_ZONE %in% modcats] <- "moderate risk"
floodarea$zones[floodarea$FLD_ZONE %in% lowcats] <- "low risk"
#table(floodarea$zones, floodarea$FLD_ZONE)

#intersecting land use and flood area
luflood <- intersect(landuse, floodarea)
```

This initial plot gives us a general idea of the distribution of the flood zones within the two counties. We can see that most of the two counties is in a FEMA-designated flood zone. However, this initial map gives us little insight to what types of land are in a flood zone, and where the risk-levels are. We are also unable to quantify the percentage of each land use type in a flood zone from this initial map.

```{r}
# getting base map
bmap <- get_tiles(luflood, crop = TRUE,provider = "CartoDB.Positron")
#plotting overall land use/ flood zones
par(mar = c(2,2,6,2))
plot(landuse,main = "Flood Hazard Areas in Sacramento and Yolo County",
     axes = FALSE)
plot_tiles(bmap,adjust = TRUE,add = TRUE)
lines(landuse)
lines(luflood, col = "dodgerblue")

```

## Further Analysis

In order to get a more detailed look, we created one plot with agricultural land, and one plot with urban land. To do this, we subset using logical operators. We then color-coded the different flood risks to get a better understanding of where each type of flood risk was in each land use class. To make this process more efficient, we used a for-loop to create the two plots. Within the loops, we first subset the intersected land use and flood zone data to the desired land-use class. Then we assigned each polygon a color based on its flood-risk group. We then plotted the result with a base map from the maptiles package.

```{r}
level_order <- c("high risk", "moderate risk", "low risk")

# modifying the factor levels of the fill variable
luflood$zoneord = factor(luflood$zones, levels = level_order)


pal <- wes_palette("Darjeeling1",3,type = "discrete")
pal <- c(pal[1],pal[3],pal[2])
usetypes <- unique(luflood$landuse)
for (i in 1:length(usetypes)){
  landsub <- luflood[luflood$landuse == usetypes[i]]
  copal <- pal[as.factor(landsub$zoneord)]
  plot(landuse[landuse$landuse == usetypes[i]],
     main = paste0("Flood Zones in ",usetypes[i], " Land in Yolo and Sacramento County"),
     axes = FALSE)
plot_tiles(bmap,adjust = TRUE,add = TRUE)
lines(landuse[landuse$landuse == usetypes[i]])
lines(landsub, col = copal)
legend(x="bottomleft",legend = c('high risk',"moderate risk","low risk"), col = pal, pch = 15,xpd = TRUE,cex = 0.8)}
```

After mapping the different flood zones, it is helpful to plot it in a different form to get a better understanding of their distribution. First, we created a bar plot showing the overall distribution of the types of flood zones. First we created a dataframe of the county, flood zones (including their type), flood zone area, and the land use type. Then, using the aggregate function in the stats package, we were able to sum the areas of each type of flood zone. We then created a bar plot with base graphics to show this data.

```{r}
#creating area dataframe
areadata <-data.frame(luflood$COUNTY,
                   luflood$zones,expanse(luflood,unit = 'km'),
                   luflood$landuse,luflood$zoneord)

names(areadata)<- c("county","flood.risk","area","land.use","order")
#total area datagrame
totarea <- stats::aggregate(area ~ flood.risk, areadata, sum)

level_order <- c("low risk", "moderate risk", "high risk")

# convert the flood_risk column to a factor with the desired level order
totarea$flood.risk <- factor(totarea$flood.risk, levels = level_order)
#plotting
par(mar = c(4,5,6,4))
barplot(totarea$area, names.arg = totarea$flood.risk, col = c(pal[1],pal[3],pal[2]),
        main = "Land Area Covered by\n Each Flood Risk\n Type in Yolo and Sacramento County",
        ylab = expression(area~(km^2)))
```

While this plot gives us a general idea of what types of flood risks are more abundant in the counties, it is still good to get more insight on the flood risks in each land use type across the two counties. To do this, we created a grouped, stacked bar plot using the ggplot2 package. To do this, we used the dataframe created in the previous section with information about the county, flood zones types, flood zone area, and the land use type.

```{r}
# ggplot(areadata, aes(x = flood.risk,y = area))+geom_col(position = 'dodge')
ggplot(data = areadata,
       aes(x = land.use,
           y = area,
           fill = order)) +
  geom_col(position = "stack") +
  facet_grid(~ county)+ scale_fill_manual(values = pal)+
  labs(title = "Distribution of Flood Risk in Yolo and Sacramento County\nAcross Urban and Agricultural Land",
       x = "land use type", y = "area in sq. km", fill = "flood risk type")+
  theme(axis.text.y = element_blank())
```

Using the created dataframe we can also calculate the proportion of flood risk type in each land use class. First, using logical operators, we can subset and sum the total area of flood zones within each land use type. Then, we can divide over the total area of each land use class.

```{r}
# getting total agriculture area
fl_ag <- sum(areadata$area[areadata$land.use=="Agricultural"])
agarea <- sum(expanse(landuse[landuse$landuse=="Agricultural"],unit = "km"))
# calculating proportion of flood-risk in ag
frag <- fl_ag/agarea

fl_urb <- sum(areadata$area[areadata$land.use=="Urban"])
# getting total urban area
urbarea<- sum(expanse(landuse[landuse$landuse=="Urban"],unit = "km"))
# calculating proportion
frurb <- fl_urb/urbarea
```

We can also calculate the proportion of each flood risk type in each land use class.

```{r}
#getting zone categories
zonecats <- unique(luflood$zones) 
# creating data frame
agzonal = data.frame(zonecats,rep(1,length(zonecats)))
names(agzonal) <- c("risk","area")

#calculating for each zone type, inserting in data frame
for (i in 1:length(zonecats)){
  agzonal$area[i] = sum(expanse(luflood[luflood$landuse=="Agricultural" & luflood$zones == zonecats[i]],
                     unit = "km"))
}


#calculating and creating data frame for urban land
urbzonal = data.frame(zonecats,rep(1,length(zonecats)))
names(urbzonal) <- c("risk","area")
for (i in 1:length(zonecats)){
urbzonal$area[i] = sum(expanse(luflood[luflood$landuse=="Urban" & luflood$zones == zonecats[i]],
                     unit = "km"))
}

```


## Conclusion

Overall, we found that nearly the entire agricultural and urban land in both Yolo and Sacramento counties falls within flood zones, with 99.5% and 96.3% of the land area affected, respectively. Notably, agricultural land appears to be more susceptible to flooding than urban land, with approximately 58.6% of agricultural land in high-risk flood zones, compared to only 6.8% of urban land in high-risk flood zones. In terms of moderate risk, urban lands fall into much more moderate risk zones, with 84.7% compared to 38.3% of agriculture. Lastly, both countries have significantly small low risk zones, agricultural land accounts for about 2.6% of the area while urban land accounts for about 4.9%.\
Because of the lack of quantification of how FEMA's flood zone risks were defined in the dataset, it was difficult to identify any relations between land-use and flood risks. FEMA's flood risk polygons were created for the National Flood Insurance Program, and consequently do not provide data that is extremely helpful in predicting flood events. Also, since the land use data was derived from a crop mapping dataset, it did not provide a full scope of the types of land in the areas of analysis. In the future, we plan to expand our analysis with additional counties, and eventually the entire state of California. We also aim to develop a model based on several predictor variables to show how flood risk varies by different location. Furthermore, we would like to investigate the relationship between flood vulnerability and other factors such as population, temperature, etc. To improve workflow efficiency, we could try to create more functions, particularly in the categorizing and calculating portions of our project.

## Works Cited

Huang, X., Stevenson, S., & Hall, A. D. (2020). Future warming and intensification of precipitation extremes: A "Double Whammy" Leading to Increasing Flood Risk in California. *Geophysical Research Letters*, *47*(16). <https://doi.org/10.1029/2020gl088679>\
\
Pattison, I., & Lane, S. N. (2011). The link between land-use management and fluvial flood risk. *Progress in Physical Geography: Earth and Environment*, *36*(1), 72--92. <https://doi.org/10.1177/0309133311425398>
